---
title: "ptp1b_seurat"
format: html
editor: visual
---

```{r setup}
# Setting the working directory for R chunks
knitr::opts_knit$set(root.dir = "/Users/spencerelliott/ptp1b_ad/kallisto_output")
# Cache results so huge analysis doesn't have to be rerun
knitr::opts_chunk$set(cache = TRUE)
```

## Analyzing the Feature-Barcode Matrix in R

The main output from pre-processing is a matrix quantifying the expression of each gene in each cell. The Seurat package for R includes many tools for analyzing this type of single-cell data.

This analysis will heavily follow the Guided Clustering Tutorial provided on the Seurat website: <https://satijalab.org/seurat/articles/pbmc3k_tutorial.html>

### Creating a Seurat object from the matrix

Let's start by loading in the Seurat package. Since the data is in matrix format, I will also use the Matrix package for some easier manipulation. Lastly the dplyr package is always helpful for piping commands and moving data around.

```{r}
library(Seurat)
library(Matrix)
library(dplyr)
```

The first step is reading in the matrix generated by bustools count, after setting my working directory to the output folder from kallisto. Let's look at the dimensions of this matrix:

```{r}
counts <- readMM("counts.mtx")
dim(counts)
```

That represents 372,063 possible cell barcodes, and 55,335 genes/transcripts from the reference transcriptome (a high number because this includes many noncoding).

First of all this matrix needs to be transposed, Seurat expects rows to be features and columns to be cell barcodes:

```{r}
counts <- t(counts)
```

Next I will bring in the lists of barcode and feature names:

```{r}
barcodes <- readLines("counts.barcodes.txt")
features <- readLines("counts.genes.txt")
```

The feature names are currently Ensembl IDs that need the version number removed.

```{r}
features <- sub("\\.\\d+$", "", features)
```

These lists can be assigned as names to the columns and rows of the matrix:

```{r}
rownames(counts) <- features
colnames(counts) <- barcodes
```

372k barcodes is an extremely high amount, it's likely that many of these are empty droplets that don't actually contain cells. Printing a summary of UMIs supports this conclusion:

```{r}
summary(colSums(counts))
# colSums(): Adds up the values in each column (# of UMIs for each cell)
# summary(): Computes some basic statistics (mean, median, min, max)
```

The mean is 65.4 while the median is only 1! That means at least half the barcodes only have 1 UMI, a sign of empty droplets. To get rid of these data points, I will filter down to barcodes associated with at least 500 UMIs.

```{r}
counts_filtered <- counts[, colSums(counts) >= 500]
```

Now the matrix with raw counts can be turned into Seurat's special data structure, which can store the analysis that I will perform. This step also allows for a bit of additional filtering:

```{r}
so <- CreateSeuratObject(counts = counts_filtered,
                         project = "ptp1b_ad",
                         min.cells = 3,
                         min.features = 200)
# min.cells: Drop any gene present in less than 3 cells
# min.features: Drop any cell with less than 200 unique features
```

I chose these numbers for filtering as a starting point (they are used in the walkthrough example on Seurat's website). There is also more quality control to look at before drawing any big conclusions.

### Filtering Low-Quality Cells

We need to remove low-quality cells from this scRNA-seq data. There are a few common ways to filter out low-quality and dying cells, including:

\- Low number of total UMIs.\
- Low number of unique genes/features.\
- High percentage of UMIs from mitochondrial genes. (Cytoplasmic RNA is lost first in dying cells, so mitochondria get overrepresented)

The first two are built into the cell metadata that Seurat gives us as "nCount_RNA" and "nFeature_RNA," respectively:

```{r}
head(so@meta.data, 5)
```

For mitochondrial content, I will need to create a new column alongside these two others.

Finding the mitochondrial genes is a little tricky because everything in the matrix is an Ensembl ID right now. I will have to go back to the annotation file and find out which ones are the mitochondrial genes.

```{r}
gtf <- read.table("../gencode.vM23.annotation.gtf",
                  sep = "\t",
                  comment.char = "#",
                  header = FALSE)
# sep = "\t": File is tab separated (gtf)
# comment.char = "#": Lines beginning with # are comments
# header = False: No header row in gtf file
```

Mouse mitochondrial genes are listed as being on chromosome "chrM" (column 'V1'), and genes are the type of feature that we are looking for (column 'V3'):

```{r}
mt.gtf <- gtf[gtf$V1 == "chrM" & gtf$V3 == "gene", ]
# gtf$V1 == "chrM": Mitochondrial chromosome
# gtf$V3 == "gene": Feature type is "gene"
```

Gene IDs are in the attribute field (column 'V9') alongside a bunch of other information that will need to be removed:

```{r}
attr <- mt.gtf$V9
head(attr)
```

I used AI to write a command with regex isolating the ID names we need:

```{r}
# Remove everything except the Ensembl ID
mt.genes <- sub('.*gene_id "?([^";]+)"?.*', '\\1', attr)
# Remove the version number
mt.genes <- sub("\\..*", "", mt.genes)
```

With the list of mitochondrial Ensembl IDs, we can calculate the mitochondrial percentage for each barcode.

```{r}
so[["percent.mt"]] <- PercentageFeatureSet(so, features = mt.genes)
# so[["percent.mt"]]: Add a new column for mitochondrial percentage
# PercentageFeatureSet(so, pattern = "^mt-"): For each barcode, calculate the percentage of genes starting with "mt-" (mitochondrial)

summary(so$percent.mt)
```

The summary looks like what we would expect! The mean is 3.7% mitochondrial genes.

Let's inspect all the cell metadata with this new column added:

```{r}
head(so@meta.data, 5)
# @meta.data: Contains cell-level metadata like # of features and UMIs
```

Finally, let's visualize each of these QC metrics with a violin plot:

```{r}
VlnPlot(so, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
```

- nFeatureRNA: The bulk of the violin plot sits at \~500-3500 genes. The outliers towards the top are very likely doublets, where two cells were mistakenly counted as one.

- nCountRNA: Most barcodes sit at \<10000 UMIs.

- percent.mt: The majority of cells are grouped up at \~5% mitochondrial content. A tail of outliers goes up as high as 50%.

To help choose cutoffs a FeatureScatter plot is also helpful. These plots show how the number of features and mitochondrial percentage differs based on UMI count:

```{r}
plot1 <- FeatureScatter(so, feature1 = "nCount_RNA", feature2 = "percent.mt")
# Plot # of UMIs vs mitochondrial content

plot2 <- FeatureScatter(so, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
# Plot # of UMIs vs # of features

plot1 + plot2
```

- nCount_RNA vs percent.mt: Cells with the highest mitochondrial content also have the lowest UMI counts, a result typically expected from dead/dying cells.

- nCount_RNA vs nFeature_RNA: There is 0.95 correlation between UMI count and gene count, which is pretty good! Points that are way to far above have way too many features and are doublet candidates.

Based on these plots we can filter out the outliers and make this data better for analysis. The Seurat tutorial example uses the following numbers:

- 200 \< nFeature_RNA \< 2500

- percent.mt \< 5

That data is made up of human cells with different results for the previous figures. Looking at my own plots, it looks like I can be a little more generous and include:

- 200 \< nFeature_RNA \< 5000

- percent.mt \< 15

This will keep the main body of each plot while removing almost all doublets and dead cells. Time to apply these filters to the data:

```{r}
so <- subset(so, subset = nFeature_RNA > 200 &
                           nFeature_RNA < 5000 &
                           percent.mt < 15)
```

### Normalization

Right now the data is still raw counts, so a cell with more total UMIs will likely look like it has more expression for each gene. Data is commonly normalized so gene expression is given in proportion to total UMIs:

```{r}
so <- NormalizeData(so, normalization.method = "LogNormalize", scale.factor = 10000)
# LogNormalize: Log transformation (common normalization method, makes exponential relationships linear)
# scale.factor = 10000: Divide counts by total UMIs and multiply by 10000 before transforming.
```

### Identifying Highly Variable Genes

The Seurat tutorial recommends identifying genes with the highest cell-to-cell variation in expression, and focusing on them for downstream analysis. This removes noise from the genes that are expressed the same across all cells. I will follow their steps to create a subset of the 2000 most variable genes in this dataset:

```{r}
so <- FindVariableFeatures(so, selection.method = "vst", nfeatures = 2000)
# selection.method = "vst": Variance stabilizing transformation (recommended method)
# nfeatures = 2000: Find the 2000 most variable features
```

Now the Seurat object stores this list of variable features. Taking a look at the top 10 gives a list of Ensembl IDs that we can look up:

```{r}
top10 <- head(VariableFeatures(so), 10)
top10
```

In order to tell what these genes are we need to use the annotation file once again, as a map. To start let's make a subset of all rows with the feature type we are looking for (gene):

```{r}
gene.gtf <- gtf[gtf$V3 == "gene",]
# gtf$V3 == "gene": All rows with "gene" as feature type
# Leave blank after the comma to keep all columns
```

The names and IDs of all these genes are found in the attributes column (column 9), and can be extracted with some more complicated regex commands. The version numbers also need to be removed from these Ensembl IDs as seen earlier.

This information can build a new data frame "mapping" Ensembl IDs to the actual gene names:

```{r}
gene.ids   <- sub('.*gene_id "?([^";]+)"?.*',   '\\1', gene.gtf$V9)
gene.ids <- sub("\\.\\d+$", "", gene.ids)

gene.names <- sub('.*gene_name "?([^";]+)"?.*', '\\1', gene.gtf$V9)

gene.map <- data.frame(ensembl_id = gene.ids,
                       gene_name  = gene.names)
```

Now the list of top 10 variable genes can be identified:

```{r}
gene.map[gene.map$ensembl_id %in% top10, ]
# gene.map[]: Return a subset of the gene map
# gene.map$ensembl_id %in% top10: Only return IDs that are in the top 10 list
```

Now is a good time to switch all these Ensembl IDs for their corresponding gene names in the Seurat object, so all the downstream analysis will be much easier to interpret:

```{r}
new.names <- gene.map$gene_name[match(rownames(so), gene.map$ensembl_id)]
# new.names <- gene.map$gene_name[...]: New vector containing the gene symbol at each position listed in brackets
# match(rownames(so), gene.map$ensembl_id): Position in the gene map for each gene ID in the Seurat object

rownames(so[["RNA"]]) <- make.unique(new.names)
# rownames(so[["RNA"]]): Update gene names in the Seurat object
# make.unique(new.names): For multiple transcripts from the same gene, give each row a unique name (required by R)
```

Rerunning the variable feature search so the list of top 10 features has these updated names:

```{r}
so <- FindVariableFeatures(so, selection.method = "vst", nfeatures = 2000)
top10 <- head(VariableFeatures(so), 10)
```

Now all of the features can be plotted as a VariableFeaturePlot:

```{r}
# Creating a variable feature plot
plot1 <- VariableFeaturePlot(so)

# Labeling the top 10 variably expressed genes
plot2 <- LabelPoints(plot = plot1, points = top10, repel = TRUE)
plot2
```

In this plot the x-axis shows the average (normalized) expression of each gene, while the y-axis shows how much this expression varies across cells. The black dots on the bottom are deemed not variable while all genes above a certain threshold are called variable. The most variable genes are the top 10, with labeled points!

### Scaling / PCA

The next goal is to do a PCA, which condenses all of these variable genes into a few principal components that clearly separate clusters of correlated cells. These separations should ideally reflect underlying biology, like the different cell types present with different levels of gene expression.

Similar to the normalization performed earlier, the data has to be scaled before proceeding as some genes are naturally expressed much more highly than others. For example, a widely used housekeeping gene will show much more expression across all cells versus a rarely used transcription factor. Scaling ensures that the principal components are not dominated by genes with this higher magnitude of expression.

```{r}
all.genes <- rownames(so)
# rownames(so): List of all genes in the Seurat object

so <- ScaleData(so, features = all.genes)
# ScaleData: Scales and centers all features in the data set
```

Now it's time to run the PCA with the variable features that were identified:

```{r}
so <- RunPCA(so, features = VariableFeatures(so))
```

This output gives a list of the principal components that were created, and which genes are positively or negatively associated with each.

In order to see the relationship of the PCs as a figure, we need an elbow plot to figure out which ones are most worth using:

```{r}
ElbowPlot(so)
```

The x-axis is the list of PCs from highest to lowest standard deviation, which is given on the y-axis. A higher standard deviation indicates more variance being represented by that PC.

Where there are big drops in standard deviation, we can say that the PCs are capturing meaningful variation. The "elbow" is where the curve flattens out and PCs are no longer different in a way that is meaningful. To be extra safe, I will include PC1-15 for the rest of the analysis. After these results, the standard deviation becomes extremely low.

Clustering the cells is done using two commands. FindNeighbors identifies cells with similar expression profiles across all 15 PCs and assigns "neighborhoods," connecting them together. FindClusters is what actually assigns each cell to a cluster.

```{r}
so <- FindNeighbors(so, dims = 1:15)
# dims = 1:15: Use principal components 1-15 as dimensions

so <- FindClusters(so, resolution = 0.5)
# resolution = 0.5: Controls the number of clusters, 0.5 is starting point used in the Seurat tutorial
```

All 15 dimensions (PCs) need to be condensed down into 2 dimensions to make a readable plot of the clusters. This is done with the UMAP command, and the figure is generated as a DimPlot:

```{r}
so <- RunUMAP(so, dims = 1:15)

DimPlot(so, reduction = "umap", label = TRUE)
```

This plot shows the relationship of the 15 clusters to one another. The next step is to figure out what these clusters actually are by looking at marker genes. The genes that are most highly expressed in a cluster indicate what type of cells or transcriptional states are present.

Seurat provides a command to find the marker genes. For assigning identities to clusters, we are interested in what genes are being positively expressed:

```{r}
so.markers <- FindAllMarkers(so, only.pos = TRUE)
# only.pos = TRUE: Only return positively expressed marker genes

head(so.markers)
```

The average log2 fold change ("avg_log2FC") measures how much more a gene is expressed in one cluster compared to the others. Looking through the top 5 overexpressed genes in each cluster is a good starting point:

```{r}
top5 <- so.markers %>%
  group_by(cluster) %>%
  slice_max(n = 5, order_by = avg_log2FC)
# group_by(cluster): Group the rows by cluster
# slice_max(n = 5): Take the top 5 rows
# order_by = avg_log2FC: Rank rows by highest average log2 fold change

top5
```

Visualizing some expression with a heatmap:

```{r}
DoHeatmap(so, features = top5$gene) + NoLegend()
```

### Assigning Identity to Clusters

The CellMarker website allows you to look up marker genes: <http://www.bio-bigdata.center/CellMarkerBrowse.jsp>

A good starting point for identifying the clusters is automatic annotation. These programs can compare the data from these cells to a reference dataset with known labels. The result can give a great head start on identifying clusters that can then be manually checked.

For this analysis I'm going to try automatic annotation with SingleR. This requires a new library and the celldex library containing reference datasets:

```{r}
library(SingleR)
library(celldex)
```

In this case, I need the dataset for mouse RNA-seq.

```{r}
ref <- celldex::MouseRNAseqData()
```

SingleR uses the raw UMI counts from the Seurat object. The most important thing it generates is a vector of cell type predictions:

```{r}
pred <- SingleR(test = counts, ref = ref, labels = ref$label.main)
# test = counts: Data to annotate
# ref = ref: Reference dataset
# labels = ref$label.main: Chooses column from the reference to use as labels. This can be 'main' or 'fine.'

head(pred$labels, 10)
```

Let's add these labels to the Seurat object, and then use them to plot again:

```{r}
so$SingleR.labels <- pred$labels

DimPlot(so, group.by = "SingleR.labels", reduction = "umap", label = TRUE)
```

Now the clusters are labeled with cell type predictions. I can start checking out the expression of one gene in different cell types, like this:

```{r}
FeaturePlot(so, features = "Ptpn1")
```

This is the first step to verifying these cell types! For example, cluster 12 was predicted to be neurons by SingleR. Looking at a canonical marker gene for neurons, Snap25:

```{r}
FeaturePlot(so, features = "Snap25")
```

It is most expressed in cluster 12! This is the type of analysis that will help verify these results. These canonical marker genes can come from knowledge/experience and can also be found in many online resources.
