---
title: "ptp1b_seurat"
format: html
editor: visual
---

```{r setup}
# Setting the working directory for R chunks
knitr::opts_knit$set(root.dir = "/Users/spencerelliott/ptp1b_ad/kallisto_output")
```

## Analyzing the Feature-Barcode Matrix in R

The main output from pre-processing is a matrix quantifying the expression of each gene in each cell. The Seurat package for R includes many tools for analyzing this type of single-cell data.

### Creating a Seurat object from the matrix

Let's start by loading in the Seurat package. Since the data is in matrix format, I will also use the Matrix package for some easier manipulation.

```{r}
library(Seurat)
library(Matrix)
```

The first step is reading in the matrix generated by bustools count, after setting my working directory to the output folder from kallisto. Let's look at the dimensions of this matrix:

```{r}
counts <- readMM("counts.mtx")
dim(counts)
```

That represents 372,063 possible cell barcodes, and 55,335 genes/transcripts from the reference transcriptome (a high number because a ton of non-coding ones are included).

First of all this matrix needs to be transposed, Seurat expects rows to be features and columns to be cell barcodes:

```{r}
counts <- t(counts)
```

Next I will bring in the lists of barcode and feature names:

```{r}
barcodes <- readLines("counts.barcodes.txt")
features <- readLines("counts.genes.txt")
```

The feature names are currently Ensembl IDs that need the version number removed.

```{r}
features <- sub("\\.\\d+$", "", features)
```

These lists can be assigned as names to the columns and rows of the matrix:

```{r}
rownames(counts) <- features
colnames(counts) <- barcodes
```

372k barcodes is an extremely high amount, it's likely that many of these are empty droplets that don't actually contain cells. Printing a summary of UMIs supports this conclusion:

```{r}
summary(colSums(counts))
# colSums(): Adds up the values in each column (# of UMIs for each cell)
# summary(): Computes some basic statistics (mean, median, min, max)
```

The mean is 65.4, while the median is only 1! That means at least half the barcodes only have 1 UMI, which suggests empty droplets. To get rid of these data points, I will filter down to barcodes associated with at least 500 UMIs.

```{r}
counts_filtered <- counts[, colSums(counts) >= 500]
```

Now the matrix with raw counts can be turned into Seurat's special data structure, which can also contain the analysis that we will perform. This step also allows for a bit of additional filtering:

```{r}
so <- CreateSeuratObject(counts = counts_filtered,
                         project = "ptp1b_ad",
                         min.cells = 3,
                         min.features = 200)
# min.cells: Drop any gene present in less than 3 cells
# min.features: Drop any cell with less than 200 unique features
```

I chose these numbers for filtering as a starting point (they are used in the walkthrough example on Seurat's website). There is also more quality control to look at before drawing any conclusions from this data.

### Including Mitochondrial Content

We need to remove low-quality cells from this scRNA-seq data. Low-quality and dying cells can often be identified by their high mitochondrial content, so the first step is a new column tracking this value for each cell.

However, finding the mitochondrial genes is a little tricky because everything in the matrix is an Ensembl ID right now. Going back to the annotation file will make this easier:

```{r}
gtf <- read.table("../gencode.vM23.annotation.gtf",
                  sep = "\t",
                  comment.char = "#",
                  header = FALSE)
```

Mouse mitochondrial genes are on chromosome "chrM":

```{r}
mt.gtf <- gtf[gtf$V1 == "chrM" & gtf$V3 == "gene", ]
```

Extract the attribute field, gene ID, and remove version numbers:

```{r}
attr <- mt.gtf$V9

mt.genes <- sub('.*gene_id "?([^";]+)"?.*', '\\1', attr)

mt.genes <- sub("\\..*", "", mt.genes)
```

Now we can calculate mitochondrial percentage.

```{r}
so[["percent.mt"]] <- PercentageFeatureSet(so, features = mt.genes)
# so[["percent.mt"]]: Add a new column for mitochondrial percentage
# PercentageFeatureSet(so, pattern = "^mt-"): For each barcode, calculate the percentage of genes starting with "mt-" (mitochondrial)
```

Summarizing this:

```{r}
summary(so$percent.mt)
```

Let's inspect all the cell metadata:

```{r}
head(so@meta.data, 5)
# @meta.data: Contains cell-level metadata like # of features and UMIs
```

Let's extract some of this data about features and counts and percents, to create a violin plot for them.

```{r}
# Not working?
# VlnPlot(so, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)

library(tidyr)
library(ggplot2)

df <- FetchData(so, vars = c("nFeature_RNA", "nCount_RNA", "percent.mt"))

df_long <- pivot_longer(
  df,
  cols = c(nFeature_RNA, nCount_RNA, percent.mt),
  names_to = "metric",
  values_to = "value"
)

ggplot(df_long, aes(x = "", y = value)) +
  geom_violin(fill = "grey80") +
  facet_wrap(~metric, scales = "free_y", ncol = 3) +
  theme_classic() +
  theme(
    axis.title.x = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank()
  )
```
