---
title: "ptp1b_seurat_final"
format:
  html:
    embed-resources: true
editor: visual
---

```{r setup}
library(here)
# Setting the random seed to replicate results
seed <- readRDS(here("results/original_seed.rds"))
assign(".Random.seed", seed, envir = .GlobalEnv)
# Cache results so huge analysis doesn't have to be rerun
knitr::opts_chunk$set(cache = TRUE, cache.lazy = FALSE)
```

## Seurat Analysis for All Samples

Running CellRanger resulted in a feature-barcode matrix for all four samples, and CellBender cleaned up these results extensively. Now I can bring my data into Seurat and perform QC and clustering analysis, like I did for just one sample earlier. I'm also going to do some more detailed preprocessing, incorporating other tools that were used on this data in the paper.

This will open the door to comparing expression in different cell types across the two treatment groups!

Once again this analysis will heavily follow the Seurat tutorial found [here](https://satijalab.org/seurat/articles/pbmc3k_tutorial.html).\
Also remember that for many of these steps I wrote up a more detailed explanation the first time around [(link)](https://spencer-elliott.github.io/ptp1b-ad/notebooks/ptp1b_seurat.html), when I was working on just one sample.

### Creating the merged Seurat object

Loading in the same packages as earlier:

```{r}
library(Seurat)
library(Matrix)
library(dplyr)
library(ggplot2)
library(here)
```

The first step is reading in the matrix. CellBender outputs the filtered matrices in a .h5 format that can be read in with Read10X_h5. This also contains all the other needed info, like lists of barcodes and features, that I needed to bring in separately before.

Seurat has a built-in command for reading in the matrix as these three files together. This will help me skip past steps I had to to before like replacing the Ensembl IDs with gene names!

```{r}
KO1 <- Read10X_h5(here("results/cellbender_output/KO_SRR36691286_cellbender_filtered_seurat.h5"))
KO2 <- Read10X_h5(here("results/cellbender_output/KO_SRR36691287_cellbender_filtered_seurat.h5"))
AD1 <- Read10X_h5(here("results/cellbender_output/AD_SRR36691288_cellbender_filtered_seurat.h5"))
AD2 <- Read10X_h5(here("results/cellbender_output/AD_SRR36691289_cellbender_filtered_seurat.h5"))
```

Compared to my Kallisto output, this matrix is also already filtered for empty droplets thanks to CellRanger and CellBender. I can go right ahead and create a Seurat object for each sample, then merge them together to perform analysis on all of them at once. I'll use separate project names for the two treatment groups to keep track of them after merging.

```{r}
KO1 <- CreateSeuratObject(counts = KO1, project = "KO", min.cells = 3, min.features = 200)
KO2 <- CreateSeuratObject(counts = KO2, project = "KO", min.cells = 3, min.features = 200)
AD1 <- CreateSeuratObject(counts = AD1, project = "AD", min.cells = 3, min.features = 200)
AD2 <- CreateSeuratObject(counts = AD2, project = "AD", min.cells = 3, min.features = 200)

so <- merge(KO1, y = list(KO2, AD1, AD2), 
            add.cell.ids = c("KO1", "KO2", "AD1", "AD2"))
```

### Filtering Low-Quality Cells

Once again Seurat provides "nCount_RNA" and "nFeature_RNA" for identifying the cells with dangerously low amounts of UMIs/features. Tracking mitochondrial percentage is also much easier this time, as CellRanger follows the naming convention where mouse mitochondrial genes start with "mt-"!

```{r}
so[["percent.mt"]] <- PercentageFeatureSet(so, pattern = "^mt-")
```

Looking at some metadata for QC, the percentages look around what we would expect:

```{r}
head(so@meta.data, 10)
```

I'll use a violin plot again to help decide cutoffs for these QC metrics:

```{r}
VlnPlot(so, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"))
```

- nFeatureRNA: The bulk of the violin plot sits at \~500-5000 genes for the KO group, and more like \~500-6000 for AD. Once again there are some outliers towards the top that are most likely doublets.

- nCountRNA: For the two groups, most barcodes sit at \<30000 UMIs.

- percent.mt: The majority of cells are grouped up below 20% mitochondrial content. A tail of outliers goes up even as high as 80%! Those are almost certainly dying cells.

Checking out some feature scatter plots will help as well:

```{r}
# Plot # of UMIs vs mitochondrial content
plot1 <- FeatureScatter(so, feature1 = "nCount_RNA", feature2 = "percent.mt")

# Plot # of UMIs vs # of features
plot2 <- FeatureScatter(so, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")

plot1 + plot2
```

- nCount_RNA vs percent.mt: Cells with the highest mitochondrial content also have the lowest UMI counts, a result typically expected from dead/dying cells.

- nCount_RNA vs nFeature_RNA: A 0.94 correlation between UMI count and gene count. Points that are way to far above have way too many features and are doublet candidates.

Since it seems to line up with what I'm seeing, I'm going to use the same filters that were used in the paper for this step:

- 500 \< nFeature_RNA \< 6000

- 1000 \< nCount_RNA \< 30000

- percent.mt \< 10

Applying this to the data:

```{r}
so <- subset(so, subset = nFeature_RNA > 500 & 
                       nFeature_RNA < 6000 & 
                       nCount_RNA > 1000 &
                       nCount_RNA < 30000 & 
                       percent.mt < 10)
```

### Normalization, Variable Genes, and Scaling

Continuing to follow the paper I'm going to use SCTransform this time, a tool that's included in the Seurat package. This performs normalization, identifies variable features, and scales the data in one easy step!

```{r}
so <- SCTransform(so, vars.to.regress = "percent.mt")
```

### PCA and Batch Correction

Now I can run the PCA on this list of variable features:

```{r}
so <- RunPCA(so)
```

One more step from the real analysis that I'm gonna insert here is batch correction with a tool called Harmony. This prevents separation in the PCs from technical errors or differences between the samples, and makes sure that the clustering is purely based on cell type. Running this and then checking out the elbow plot of PCs:

```{r}
library(harmony)

so <- RunHarmony(so, group.by.vars = "orig.ident", assay.use = "SCT")

ElbowPlot(so)
```

Similar to my previous analysis, the curve really starts to flatten out around PC 15 or even a little earlier. However I'm going to stick with the paper where they keep PCs 1-30.

```{r}
so <- FindNeighbors(so, dims = 1:30, reduction = "harmony")
so <- FindClusters(so, resolution = 0.15)
```

Now I can run UMAP to condense these PCs down to 2 dimensions, and make a readable map with the clusters.

```{r}
so <- RunUMAP(so, dims = 1:30, reduction = "harmony")
DimPlot(so, reduction = "umap", label = TRUE)
```

This looks pretty good again, with many distinct clusters to check out. The next step is annotating each cluster with a cell type based on which marker genes it is expressing.

### Assigning Identity to Clusters

I'll try running an automatic annotation again with SingleR as a starting point. Loading packages and the reference dataset for mouse RNA-seq:

```{r}
library(SingleR)
library(celldex)

ref <- celldex::MouseRNAseqData()
```

Now I'll take the raw UMI counts from this Seurat object and pass them over to SingleR, which will create a vector with cell type predictions for each cluster:

```{r}
counts.singleR <- GetAssayData(so, layer = "counts")

pred <- SingleR(test = counts.singleR, ref = ref, labels = ref$label.main)

head(pred$labels, 10)

```

Adding these labels to the Seurat object, and using them to plot again:

```{r}
so$SingleR.labels <- pred$labels

# Plot with SingleR labels
DimPlot(so, group.by = "SingleR.labels", reduction = "umap", label = TRUE)
```

These labels will give a first idea of what each of the clusters is representing. But the next step is a manual annotation based on canonical marker genes. Luckily, a list of marker genes to use can be found in Figure 4B of the supplemental information, a heatmap with marker genes that are overexpressed in each cluster!

For example, I'll look at the marker genes from a cluster the paper identified as astrocytes. The heatmap shows this cluster overexpressing the marker genes Aqp4, Acsbg1, and Slc1a2.

```{r}
FeaturePlot(so, features = "Aqp4")
FeaturePlot(so, features = "Acsbg1")
FeaturePlot(so, features = "Slc1a2")
```

It looks like these marker genes are all clearly overexpressed in cluster 3. It will be fair to go ahead and label this clusters as my astrocytes.

Now I can manually go through every gene from the heatmap and locate each of the same cell populations found in the paper's UMAP (Supplemental info Fig.4A). It is a pretty long process and many of the in-between clusters will be left to my own judgment.

It's important to note that a lot of these genes are overexpressed in a certain cell type but also present elsewhere. For example the gene Ttr is a marker gene of epithelial cells (choroid plexus cells). Taking a look at its expression along with the other epithelial markers:

```{r}
FeaturePlot(so, features = "Ttr", label = TRUE)
FeaturePlot(so, features = "Ecrg4", label = TRUE)
FeaturePlot(so, features = "Enpp2", label = TRUE)
```

It is overexpressed in the cluster labeled as epithelial cells, where the other epithelial cell markers are also present. But it's also present in a few random places around the other cell types. This is normal and doesn't mean those are epithelial cells too.

The ones that I'm really uncertain of, I will leave as "Unknown." When needed I'll use the automatic annotations as backup. With this in mind let's give it a try:

```{r}
# Astrocyte: Aqp4, Acsbg1, Slc1a2
FeaturePlot(so, features = "Aqp4", label = TRUE)
FeaturePlot(so, features = "Acsbg1", label = TRUE)
FeaturePlot(so, features = "Slc1a2", label = TRUE)
# Cluster 3

# Choroid plexus cell: Ttr, Ecrg4, Enpp2
FeaturePlot(so, features = "Ttr", label = TRUE)
FeaturePlot(so, features = "Ecrg4", label = TRUE)
FeaturePlot(so, features = "Enpp2", label = TRUE)
# Clusters 2, 4, 9, 12, 13, 15

# Endothelial cell: Flt1, Slco1a4, Mecom
FeaturePlot(so, features = "Flt1", label = TRUE)
FeaturePlot(so, features = "Slco1a4", label = TRUE)
FeaturePlot(so, features = "Mecom", label = TRUE)
# Cluster 14

# Macrophage: F13a1, H2-Aa, Lyve1
FeaturePlot(so, features = "F13a1", label = TRUE)
FeaturePlot(so, features = "H2-Aa", label = TRUE)
FeaturePlot(so, features = "Lyve1", label = TRUE)
# Cluster 11

# Microglia: P2ry12, Hexb, Trem2
FeaturePlot(so, features = "P2ry12", label = TRUE)
FeaturePlot(so, features = "Hexb", label = TRUE)
FeaturePlot(so, features = "Trem2", label = TRUE)
# Clusters 0, 7

# Neuron: Syt1, Snap25, Nrn1
FeaturePlot(so, features = "Syt1", label = TRUE)
FeaturePlot(so, features = "Snap25", label = TRUE)
FeaturePlot(so, features = "Nrn1", label = TRUE)
# Cluster 10

# Oligodendrocyte: Ermn, Mog, Aspa
FeaturePlot(so, features = "Ermn", label = TRUE)
FeaturePlot(so, features = "Mog", label = TRUE)
FeaturePlot(so, features = "Aspa", label = TRUE)
# Clusters 1, 5, 16

# OPC: Pdgfra, Cacng4, Vcan
FeaturePlot(so, features = "Pdgfra", label = TRUE)
FeaturePlot(so, features = "Cacng4", label = TRUE)
FeaturePlot(so, features = "Vcan", label = TRUE)
# Clusters 8

# T cell: Cd3d, Cd3e, Cd3g
FeaturePlot(so, features = "Cd3d", label = TRUE)
FeaturePlot(so, features = "Cd3e", label = TRUE)
FeaturePlot(so, features = "Cd3g", label = TRUE)
# Clusters 6

# Other / too uncertain: Clusters 17, 18
```

Now I'll create a vector of these manual cluster annotations:

```{r}
cluster_annotations <- c(
  "0" = "Microglia",
  "1" = "Oligodendrocyte",
  "2" = "Choroid plexus cell",
  "3" = "Astrocyte",
  "4" = "Choroid plexus cell",
  "5" = "Oligodendrocyte",
  "6" = "T cell",
  "7" = "Microglia",
  "8" = "OPC",
  "9" = "Choroid plexus cell",
  "10" = "Neuron",
  "11" = "Macrophage",
  "12" = "Choroid plexus cell",
  "13" = "Choroid plexus cell",
  "14" = "Endothelial cell",
  "15" = "Choroid plexus cell",
  "16" = "Oligodendrocyte",
  "17" = "Unknown",
  "18" = "Unknown"
)
```

Time to plot with these new cell types, and save the plot as a file:

```{r}
# Creating a column with the identity of each cell under manual annotations
so@meta.data$cell_type <- cluster_annotations[as.character(so$seurat_clusters)]
# so@meta.data$cell_type <-: Create a new column in Seurat metadata called cell_type
# as.character(so$seurat_clusters): Get cluster number for each cell
# cluster_annotations[]: Match each cluster number with its label in the new vector

# Use cell_type as the new identity
Idents(so) <- "cell_type"
# Idents(): Where cell idenities are stored in the Seurat object

# Remove uncertain clusters (17+18) and plot with new labels
so <- subset(so, cell_type != "Unknown")
DimPlot(so, reduction = "umap", label = TRUE)
```

Now I can compare the expression of one gene across treatment groups, and see what populations it is in! This will be the foundation for some other analyses I want to try out.

## Ptpn1 Expression

Working through the description of RNA-seq analysis:

> *"After quality control and filtering, a total of 31,740 cells across two genotypes were arranged by uniform manifold approximation and projection (UMAP) for visualization."*

Double checking how many cells I have been using:

```{r}
ncol(so)
```

Pretty close, maybe their filters are a little more strict. Maybe they also got rid of some of the trickier clusters, like that fibroblasts one. Let's continue:

> *"Clusters were then manually annotated as astrocytes, choroid plexus cells, endothelial cells, macrophages, microglia, neurons, oligodendrocytes, OPCs, and T cells, based on the expression of cell type–specific signature genes."*

That's what I did in the last step! Next:

> *"Then, we investigated the expression level of PTP1B across all cell types in the brain. Interestingly, Ptpn1, the gene that encodes PTP1B, was found to be highly expressed in immune cells, including microglia, macrophages, and T cells (Fig. 3B), suggesting a potential role in brain immune regulation."*

Time to see where Ptpn1 is expressed in my samples. It had better at least be different between the two treatment groups because it's knocked out. But let's also see which cell types show the greatest difference:

```{r}
FeaturePlot(so, features = "Ptpn1", split.by = "orig.ident", label = TRUE)
```

This feature plot shows the Ptpn1 knockout very clearly, as well as higher expression for the AD group in immune cells (microglia and macrophage, mainly).

These results are displayed in the paper as a violin plot with Ptpn1 expression in the various cell types for AD cells. To recreate this figure for comparison, the first thing to do is create a subset with this group, and reassign the manual cell type annotations.

```{r}
so_ad <- subset(so, subset = orig.ident == "AD")
Idents(so_ad) <- "cell_type"
```

First I have to create a list of cell types in the same order as the paper's figure (to make comparison easier) and remove any cell types that the paper was not interested in:

```{r}
# Vector of the cell type names in order
cell_order <- c("Choroid plexus cell", "Neuron", "Microglia", "Oligodendrocyte", "Astrocyte", "T cell", "OPC", "Macrophage", "Endothelial cell")

# Remove all cell types that aren't in the paper's analysis
so_ad <- subset(so_ad, cell_type %in% cell_order)

# Set the cell type names
so_ad$cell_type <- factor(so_ad$cell_type, levels = cell_order)
# factor(): Group the data into fixed categories
# levels = cell_order: List of categories to use (in order)
```

Now I can create the violin plot:

```{r}
VlnPlot(so_ad, features = "Ptpn1", pt.size = 0) +
  labs(x = "Cell Types", y = "Ptpn1 Expression") +
  theme(legend.position = "none") +
  stat_summary(fun = median, geom = "point")
# pt.size = 0: Don't show individual data points
# stat_summary(): Compute and plot a statistic for each group
# fun = median: Plot the median for each group
```

It looks very similar to the actual results! Microglia and macrophage have the highest levels of Ptpn1 expression, with microglia slightly higher. T cells are the next behind. This single cell data shows Ptpn1 being highly expressed in immune cells.

## Microglia Analysis

The microglia are the brain's primary immune cells, they play the most crucial role in Alzheimer's disease and are the focus of this paper. To continue the analysis they split the microglia into several subtypes and checked out the differences in gene expression there.

To follow in these footsteps, I'll create a subset of the data containing only cells that were annotated as microglia, and then redo the analysis from above.

```{r}
# New subset of microglia
microglia <- subset(so, idents = "Microglia")

# Normalization
microglia <- SCTransform(microglia, vars.to.regress = "percent.mt")

# PCA
microglia <- RunPCA(microglia, assay = "SCT")

# Batch correction
microglia <- RunHarmony(microglia, group.by.vars = "orig.ident", assay.use = "SCT")

# Elbow plot
ElbowPlot(microglia)
```

Small intermission here - it looks like 10 PCs is best for the elbow plot this time.

```{r}
# Clustering the cells
microglia <- FindNeighbors(microglia, dims = 1:10)
microglia <- FindClusters(microglia, resolution = 0.5)

# Running UMAP
microglia <- RunUMAP(microglia, dims = 1:10)

# Plot
DimPlot(microglia, reduction = "umap", label = TRUE)
```

Compared to the actual result there is way too much going on in this plot. Since this is a smaller population of cells, it makes sense to try a lower resolution for creating the clusters.

```{r}
# Clustering the cells - lower resolution
microglia <- FindClusters(microglia, resolution = 0.1)

# Running UMAP
microglia <- RunUMAP(microglia, dims = 1:10)

# Plot
DimPlot(microglia, reduction = "umap", label = TRUE)
```

That looks much better! But there is still a rogue cluster way to the left. It's likely that this is other cells that ended up annotated as microglia or in the same cluster. Let's see their genes:

```{r}
DefaultAssay(microglia) <- "SCT"
microglia <- PrepSCTFindMarkers(microglia)
cluster3.markers <- FindMarkers(microglia, ident.1 = "3")
head(cluster3.markers, 20)
```

None of the markers I used for microglia are actually listed here, but there is Ecrg4 which was a marker for choroid plexus cells. It really looks like this cluster just happened to get mixed in. Checking the other rogue cluster:

```{r}
cluster4.markers <- FindMarkers(microglia, ident.1 = "4")
head(cluster4.markers, 20)
```

It also contains none of the microglia marker genes. I will go ahead and remove these, which is likely what was done in the paper as well.

```{r}
microglia <- subset(microglia, idents = c("0", "1", "2"))
DimPlot(microglia, reduction = "umap", label = TRUE)
```

And there's one cell that's a massive outlier on the UMAP. It could be a damaged cell, or caught in a transitional state or something like that. It's compressing the whole rest of my plot, so I will need to remove that as well.

```{r}
microglia <- subset(microglia, subset = umap_1 > -10)
DimPlot(microglia, reduction = "umap", label = TRUE)
```

New clusters look good!

> This revealed three distinct subclusters: homeostatic microglia (Homeostatic; expressing markers: Fcrls, Tmem119 and P2ry12), disease-associated microglia (DAM; expressing markers: Axl, Ctsl, and Trem2) and interferon-responsive microglia (IFN; expressing markers: Oasl2, Stat1, and Irf7).

Figuring out which clusters correspond to each of these subpopulations will be similar to my manual annotations for all cell types:

```{r}
# Homeostatic microglia: Fcrls, Tmem119, P2ry12
FeaturePlot(microglia, features = "Fcrls", label = TRUE)
FeaturePlot(microglia, features = "Tmem119", label = TRUE)
FeaturePlot(microglia, features = "P2ry12", label = TRUE)
# Cluster 0

# Disease-associated microglia: Axl, Ctsl, Trem2
FeaturePlot(microglia, features = "Axl", label = TRUE)
FeaturePlot(microglia, features = "Ctsl", label = TRUE)
FeaturePlot(microglia, features = "Trem2", label = TRUE)
# Cluster 1

# Interferon-responsive microglia: Oasl2, Stat1, and Irf7
FeaturePlot(microglia, features = "Oasl2", label = TRUE)
FeaturePlot(microglia, features = "Stat1", label = TRUE)
FeaturePlot(microglia, features = "Irf7", label = TRUE)
# Cluster 2
```

Labeling the clusters based on this investigation, and plotting the result:

```{r}
microglia_annotations <- c(
  "0" = "Homeostatic",
  "1" = "DAM",
  "2" = "IFN"
)
microglia@meta.data$cell_type <- microglia_annotations[as.character(microglia$seurat_clusters)]
Idents(microglia) <- "cell_type"

DimPlot(microglia, reduction = "umap", label = TRUE)
```

Now I have clustered the microglia and can compare gene expression between these populations!
